Structured Versus Decorated Cospans

### 2020-11-16 cospans.tex
a few edits

\author{John\ C.\ Baez$^{1,2}$, Kenny Courser$^1$, and Christina Vasilakopoulou$^3$}
\address{$^1$Department of Mathematics, University of California, Riverside CA, USA 92521}
\address{$^2$Centre for Quantum Technologies, National University of Singapore, Singapore 117543}
\address{$^3$Department of Mathematics, University of Patras, Greece 265 04}
\email{baez@math.ucr.edu, kcour001@ucr.edu, cvasilak@math.upatrs.gr}

@@

\begin{thm} \label{thm:equiv}
Suppose $\ca{A}$ has finite colimits and $(F,\phi,\phi_0)\maps(\ca{A},+) \to (\bicat{Cat},\times,1)$ is a symmetric lax monoidal pseudofunctor that factors through $\bicat{Rex}$ as an ordinary pseudofunctor. 
Then the symmetric monoidal double category $F\lCsp$ of decorated cospans is equivalent to the symmetric monoidal double category $_L\lCsp(\inta F)$ of structured cospans, where $L \maps \A \to \inta F$ is the left adjoint of the induced Grothendieck opfibration $\U\maps \inta F \to \A$.
\end{thm}

The conditions of the above theorem relate to the existence of left adjoints for opfibrations; we first sketch this underlying framework, and then we proceed to the proof of the theorem. The basics of fibration theory needed for our purposes are recalled in \cref{sec:fibrations}.

@@

Ongoing work \cite{CV} includes a related discussion as well as a result in the opposite direction.
%For a discussion on the unusual strict cocontinuity condition...
%\begin{lem}\label{prop:CV}\cite[Proposition~3.3]{CV}
%Suppose $\U\maps\X\to\A$ is a right adjoint `left inverse'. If $\X$ and $\A$ have chosen pushouts and initial objects and $\U$ strictly preserves them, %then $\U$ is an opfibration.
%\end{lem}

@@

@@

\begin{proof}[Proof of \cref{thm:equiv}]
Recall from \cref{thm:decorated_cospans} that the double category of decorated cospans $\dcsp{F}$ has objects and vertical 1-morphisms as in $\A$, while a horizontal 1-cell $a\tickar b$ is cospan $\cspn{a}{m}{b}{}{}$ in $\A$ decorated by an object $x\in F(m)$ and a 2-morphism is a map $k\maps m\to m'$ together with a morphism $(Fk)(x)\to x'$.

By \cref{cor:fcocMonGroth}, when $F$ as an ordinary pseudofunctor factors through $\bicat{Rex}$, the (monoidal) Gro\-the\-ndieck construction gives rise to a finitely cocomplete $\inta F$ such that the corresponding opfibration $U_F\maps(\inta{F},+)\to(\A,+)$ preserves all finite colimits. %Christina: maybe the cocartesian structures are redundant to write here, since U_F really preserves all finite colimits anyways.
Since in particular it preserves the initial object, \cref{prop:opfibtolari} through \cref{lem:fibrewiselimits} applies 
%which equivalently means that all fibres have initial objects and the reindexing functors preserve them by , %Christina: there is a little bit of circularity here. We knew fibres have and reindexing functors preserve from the factorization already! Change Gray accordingly perhaps?
to construct a left adjoint $L_F\maps\A\to\inta F$ which is also a right inverse, namely $U_FL_F=\mathrm{id}_\A$. Explicitly, this left adjoint is given by $L(a)=(a,\bot_a)$, picking the initial object in the finitely cocomplete fibre also expressed as $F(u_a)\circ\phi_0(*)$ in \cref{eq:initialinfibre}. Diagrammatically,
\begin{displaymath}
 F\maps\X\to\bicat{Cat}\quad\mapsto\quad\begin{tikzcd}[baseline=.3]\inta F\ar[d,"U_F"'] \\ \X \end{tikzcd}\quad\mapsto\quad\begin{tikzcd}\X\ar[r,bend left,pos=.55,"L_F"]\ar[r,phantom,"\bot"description] & \inta F\ar[l,bend left,pos=.45,"U_F"]\end{tikzcd}
\end{displaymath}
roughly describes the processes between the original $F$ and the resulting $L_F$.

@@

\subsection{2-categories}\label{sec:2cats}
For standard 2-categorical material, we refer the reader to e.g. \cite{KS} or \cite{DS} for monoidal structures.
Recall that a \define{pseudofunctor} $F\maps\bicat{A}\to\bicat{B}$ between bicategories $\bicat{A}$ and $\bicat{B}$ is functorial up to coherent natural isomorphism, namely for composable arrows we have $F(g\circ f)\cong Fg\circ Ff$ and $F(1_a)\cong1_{Fa}$ satisfying standard axioms. 
Given pseudofunctors $F,G\maps \bicat{A} \to \bicat{B}$, a \define{pseudonatural transformation} $\sigma$ consists of
\begin{itemize}
\item for each object $a \in \bicat{A}$, a morphism $\sigma_a \maps F(a) \to G(a)$ in $\bicat{B}$ 
\item 
%for every pair of objects $a$ and $b$ of $\mathbf{A}$, we have natural isomorphisms
%\[
%\begin{tikzpicture}[scale=1.5]
%\node (A) at (0,0) {$\mathbf{A}(a,b)$};
%\node (B) at (2,0) {$\mathbf{X}(F(a),F(b))$};
%\node (C) at (0,-1) {$\mathbf{X}(G(a),G(b))$};
%\node (D) at (2,-1) {$\mathbf{X}(F(a),G(b))$};
%\node (E) at (1,-0.5) {$\sigma_{a,b} \Nearrow$};
%\path[->,font=\scriptsize,>=angle 90]
%(A) edge node[above]{$F$} (B)
%(B) edge node[right]{$(\sigma_{b})_*$} (D)
%(A) edge node[left]{$G$} (C)
%(C) edge node[above]{$(\sigma_a)^*$} (D);
%\end{tikzpicture}
%\]
%where $(\sigma_a)^*$ and $(\sigma_b)_*$ are the functors induced by precomposition and postcompositon, respectively. Thus 
for each morphism $f \maps a \to b$ in $\mathbf{A}$, an invertible natural 2-morphism $\sigma_f \maps G(f) \sigma_a \xrightarrow{\sim} \sigma_{b} F(f)$ in $\mathbf{B}$
%:
%\[
%\begin{tikzpicture}[scale=1.5]
%\node (A) at (0,0) {$F(a)$};
%\node (B) at (1,0) {$F(b)$};
%\node (C) at (0,-1) {$G(a)$};
%\node (D) at (1,-1) {$G(b)$};
%\node (E) at (0.5,-0.5) {$\sigma_f \Nearrow$};
%\path[->,font=\scriptsize,>=angle 90]
%(A) edge node[above]{$F(f)$} (B)
%(B) edge node[right]{$\sigma_{b}$} (D)
%(A) edge node[left]{$\sigma_a$} (C)
%(C) edge node[above]{$G(f)$} (D);
%\end{tikzpicture}
%\]
compatible with composition and identities.
\end{itemize}
%such that for each composable pair of morphisms $f \maps a \to b$ and $g \maps b \to c$ of $\mathbf{A}$, the following diagrams commute:
%\[
%\begin{tikzpicture}[scale=1.5]
%\node (A) at (0,0.5) {$(G(g)G(f))\sigma_a$};
%\node (A') at (2,0.5) {$G(g)(G(f) \sigma_a)$};
%\node (B) at (0,-0.5) {$G(gf) \sigma_a$};
%\node (C) at (4,0.5) {$G(g) (\sigma_b F(f))$};
%\node (C') at (4,-0.5) {$\sigma_c F(gf)$};
%\node (D) at (6,0.5) {$(G(g) \sigma_b) F(f)$};
%\node (D') at (8,-0.5) {$\sigma_c (F(g)F(f))$};
%\node (F) at (8,0.5) {$(\sigma_c F(g)) F(f)$};
%\path[->,font=\scriptsize,>=angle 90]
%(A) edge node[above]{$a'$} (A')
%(A) edge node[left]{$\psi 1_{\sigma_a}$} (B)
%(A') edge node[above]{$1_{G(g)} \sigma_f$} (C)
%(B) edge node[above]{$\sigma_{gf}$} (C')
%(C) edge node [above] {${a'}^{-1}$} (D)
%(D') edge node [above] {$1_{\sigma_c} \phi$} (C')
%(D) edge node [above] {$\sigma_g 1_{F(f)}$} (F)
%(F) edge node [right] {$a'$} (D');
%\end{tikzpicture}
%\]
%\[
%\begin{tikzpicture}[scale=1.5]
%\node (A) at (0,0) {$1_{G(a)} \sigma_a$};
%\node (B) at (1,1) {$G(1_a) \sigma_a$};
%\node (C) at (2,0) {$\sigma_a F(1_a)$};
%\node (D) at (0.5,-1) {$\sigma_a$};
%\node (E) at (1.5,-1) {$\sigma_a 1_{F(a)}$};
%\path[->,font=\scriptsize,>=angle 90]
%(A) edge node[left]{$\psi 1_{\sigma_a}$} (B)
%(B) edge node[right]{$\sigma_{1_a}$} (C)
%(A) edge node [left] {$\ell'$} (D)
%(D) edge node [above] {${\rho'}^{-1}$} (E)
%(E) edge node [right] {$1_{\sigma_a} \phi$} (C);
%\end{tikzpicture}
%\]
%Christina: I removed the above, they are not needed I believe. Check again in the end.
We denote by $[\A,\bicat{Cat}]_\pse$ the 2-category 
of pseudofunctors, pseudonatural transformations and modifications from an ordinary category $\A$ viewed as a 2-category with trivial 2-cells, into $\bicat{Cat}$. This is also referred to as the 2-category of \emph{opindexed categories}, since an indexed category is a contravariant pseudofunctor into $\bicat{Cat}$.

A \define{monoidal} bicategory $\A$ comes with a pseudofunctor $\otimes\maps\A\times\A\to\A$ and a unit object $I$ that are associative and unital up to coherent equivalence. A \define{braided} monoidal bicategory comes with a pseudonatural equivalence $\beta_{a,b}\maps a\ot b\to b\ot a$ and appropriate invertible modifications; it is \define{sylleptic} if there is an appropriate invertible modification $1_{a\ot b}\Rrightarrow\beta_{b,a}\circ\beta_{a,b}$ and it is \define{symmetric} if one further axiom holds. A \define{lax monoidal} pseudofunctor between monoidal bicategories $F\maps\bicat{A}\to\bicat{B}$ is a pseudofunctor equipped with pseudonatural transformations with components $\mu_{a,b}\maps Fa\otimes Fb\to F(a\otimes b)$ and $\mu_0\maps I\to FI$ along with coherent invertible modifications for associativity and unitality. This is commonly also called `weak monoidal' pseudofunctor. A \define{braided lax monoidal} pseudofunctor between symmetric monoidal bicategories comes with an invertible modification with components $F(\beta_{a,b})\circ \mu_{a,b}\cong\mu_{b,a}\circ \beta_{Fa,Fb}$.

>>>>

\author{John\ C.\ Baez$^{1,2}$, Kenny Courser$^1$, and Christina Vasilakopoulou$^3$}
\address{$^1$Department of Mathematics, University of California, Riverside CA, USA 92521}
\address{$^2$Centre for Quantum Technologies, National University of Singapore, Singapore 117543}
\address{$^3$Department of Mathematics, University of Patras, Greece 265 04}
\email{baez@math.ucr.edu, kcour001@ucr.edu, cvasilak@math.upatras.gr}

@@

\begin{thm} \label{thm:equiv}
Suppose $\ca{A}$ has finite colimits and $(F,\mu,\mu_0)\maps(\ca{A},+) \to (\bicat{Cat},\times)$ is a symmetric lax monoidal pseudofunctor that factors through $\bicat{Rex}$ as an ordinary pseudofunctor. 
Then the symmetric monoidal double categories $F\lCsp$ of decorated cospans and $_L\lCsp(\inta F)$ of structured cospans are equivalent, where $L \maps \A \to \inta F$ is a left adjoint of the induced Grothendieck opfibration $\U\maps \inta F \to \A$.
\end{thm}

The conditions of the above theorem relate to the existence of colimits as well as left adjoints for opfibrations. We first sketch this underlying framework, and then we proceed to the proof of the theorem. The basics of fibration theory needed for our purposes are recalled in \cref{sec:fibrations}.

@@

Work in progress \cite{CV} includes a related discussion as well as a result in the opposite direction.
%For a discussion on the unusual strict cocontinuity condition...
%\begin{lem}\label{prop:CV}\cite[Proposition~3.3]{CV}
%Suppose $\U\maps\X\to\A$ is a right adjoint `left inverse'. If $\X$ and $\A$ have chosen pushouts and initial objects and $\U$ strictly preserves them, %then $\U$ is an opfibration.
%\end{lem}

@@

\begin{proof}[Proof of \cref{thm:equiv}]
Recall from \cref{thm:decorated_cospans} that the double category of decorated cospans $\dcsp{F}$ has objects and vertical 1-morphisms as in $\A$, while horizontal 1-cells are cospans $\cspn{a}{m}{b}{}{}$ in $\A$ decorated by an $x\in F(m)$ and 2-morphisms are maps of cospans $k\maps m\to m'$ together with $\tau\maps(Fk)(x)\to x'$ in $F(m')$. %!!! Christina: Should we call them `generalized maps of cospans'? Since maps of cospans are usually the ones with fixed (co)domains

By \cref{cor:fcocMonGroth}, when $F$ as an ordinary pseudofunctor factors through $\bicat{Rex}$, the Gro\-the\-ndieck construction gives rise to a finitely cocomplete $\inta F$ such that the corresponding opfibration $U\maps\inta{F}\to\A$ preserves all finite colimits.
Since in particular it preserves the initial object, \cref{lem:fibrewiselimits,prop:opfibtolari} apply 
%which equivalently means that all fibres have initial objects and the reindexing functors preserve them by , %Christina: there is a little bit of circularity here. We knew fibres have and reindexing functors preserve from the factorization already! Change Gray accordingly perhaps?
and produce a left adjoint $L\maps\A\to\inta F$ to $U$, with $UL=1_\A$. 
Diagrammatically,
\begin{displaymath}
 F\maps\A\to\bicat{Cat}\quad\mapsto\quad\begin{tikzcd}[baseline=.3]\inta F\ar[d,"U"'] \\ \A \end{tikzcd}\quad\mapsto\quad\begin{tikzcd}\A\ar[r,bend left,pos=.55,"L"]\ar[r,phantom,"\bot"description] & \inta F\ar[l,bend left,pos=.45,"U"]\end{tikzcd}
\end{displaymath}
describes the construction of the adjunction from the original $F$. Explicitly, the left adjoint picks the initial object in the finitely cocomplete fibre $L(a)=(a,\bot_a)$, also expressed by $F(u_a)\circ\mu_0(*)$ as in \cref{eq:initialinfibre}.

@@

@@

\subsection{2-categories}\label{sec:2cats}
For standard 2-categorical material, we refer the reader to e.g. \cite{KS} or \cite{DS} for monoidal structures. %!!!Christina: also John's paper on monoidal 2-structures?
Recall that a \define{pseudofunctor} $F\maps\bicat{A}\to\bicat{B}$ between bicategories $\bicat{A}$ and $\bicat{B}$ is functorial up to coherent natural isomorphism, namely for composable arrows we have $F(g\circ f)\cong Fg\circ Ff$ and $F(1_a)\cong1_{Fa}$ satisfying standard axioms. 
Given pseudofunctors $F,G\maps \bicat{A} \to \bicat{B}$, a \define{pseudonatural transformation} $\sigma$ consists of
\begin{itemize}
\item for each object $a \in \bicat{A}$, a morphism $\sigma_a \maps F(a) \to G(a)$ in $\bicat{B}$ 
\item 
%for every pair of objects $a$ and $b$ of $\mathbf{A}$, we have natural isomorphisms
%\[
%\begin{tikzpicture}[scale=1.5]
%\node (A) at (0,0) {$\mathbf{A}(a,b)$};
%\node (B) at (2,0) {$\mathbf{X}(F(a),F(b))$};
%\node (C) at (0,-1) {$\mathbf{X}(G(a),G(b))$};
%\node (D) at (2,-1) {$\mathbf{X}(F(a),G(b))$};
%\node (E) at (1,-0.5) {$\sigma_{a,b} \Nearrow$};
%\path[->,font=\scriptsize,>=angle 90]
%(A) edge node[above]{$F$} (B)
%(B) edge node[right]{$(\sigma_{b})_*$} (D)
%(A) edge node[left]{$G$} (C)
%(C) edge node[above]{$(\sigma_a)^*$} (D);
%\end{tikzpicture}
%\]
%where $(\sigma_a)^*$ and $(\sigma_b)_*$ are the functors induced by precomposition and postcompositon, respectively. Thus 
for each morphism $f \maps a \to b$ in $\mathbf{A}$, an invertible natural 2-morphism $\sigma_f \maps G(f) \sigma_a \xrightarrow{\sim} \sigma_{b} F(f)$ in $\mathbf{B}$
%:
%\[
%\begin{tikzpicture}[scale=1.5]
%\node (A) at (0,0) {$F(a)$};
%\node (B) at (1,0) {$F(b)$};
%\node (C) at (0,-1) {$G(a)$};
%\node (D) at (1,-1) {$G(b)$};
%\node (E) at (0.5,-0.5) {$\sigma_f \Nearrow$};
%\path[->,font=\scriptsize,>=angle 90]
%(A) edge node[above]{$F(f)$} (B)
%(B) edge node[right]{$\sigma_{b}$} (D)
%(A) edge node[left]{$\sigma_a$} (C)
%(C) edge node[above]{$G(f)$} (D);
%\end{tikzpicture}
%\]
compatible with composition and identities.
\end{itemize}
%such that for each composable pair of morphisms $f \maps a \to b$ and $g \maps b \to c$ of $\mathbf{A}$, the following diagrams commute:
%\[
%\begin{tikzpicture}[scale=1.5]
%\node (A) at (0,0.5) {$(G(g)G(f))\sigma_a$};
%\node (A') at (2,0.5) {$G(g)(G(f) \sigma_a)$};
%\node (B) at (0,-0.5) {$G(gf) \sigma_a$};
%\node (C) at (4,0.5) {$G(g) (\sigma_b F(f))$};
%\node (C') at (4,-0.5) {$\sigma_c F(gf)$};
%\node (D) at (6,0.5) {$(G(g) \sigma_b) F(f)$};
%\node (D') at (8,-0.5) {$\sigma_c (F(g)F(f))$};
%\node (F) at (8,0.5) {$(\sigma_c F(g)) F(f)$};
%\path[->,font=\scriptsize,>=angle 90]
%(A) edge node[above]{$a'$} (A')
%(A) edge node[left]{$\psi 1_{\sigma_a}$} (B)
%(A') edge node[above]{$1_{G(g)} \sigma_f$} (C)
%(B) edge node[above]{$\sigma_{gf}$} (C')
%(C) edge node [above] {${a'}^{-1}$} (D)
%(D') edge node [above] {$1_{\sigma_c} \phi$} (C')
%(D) edge node [above] {$\sigma_g 1_{F(f)}$} (F)
%(F) edge node [right] {$a'$} (D');
%\end{tikzpicture}
%\]
%\[
%\begin{tikzpicture}[scale=1.5]
%\node (A) at (0,0) {$1_{G(a)} \sigma_a$};
%\node (B) at (1,1) {$G(1_a) \sigma_a$};
%\node (C) at (2,0) {$\sigma_a F(1_a)$};
%\node (D) at (0.5,-1) {$\sigma_a$};
%\node (E) at (1.5,-1) {$\sigma_a 1_{F(a)}$};
%\path[->,font=\scriptsize,>=angle 90]
%(A) edge node[left]{$\psi 1_{\sigma_a}$} (B)
%(B) edge node[right]{$\sigma_{1_a}$} (C)
%(A) edge node [left] {$\ell'$} (D)
%(D) edge node [above] {${\rho'}^{-1}$} (E)
%(E) edge node [right] {$1_{\sigma_a} \phi$} (C);
%\end{tikzpicture}
%\]
%Christina: I removed the above, they are not needed I believe. Check again in the end.
We denote by $[\bicat{A},\bicat{B}]_\pse$ the 2-category 
of pseudofunctors, pseudonatural transformations and modifications.

A \define{monoidal} bicategory $\bicat{A}$ comes with a pseudofunctor $\otimes\maps\bicat{A}\times\bicat{A}\to\bicat{A}$ and a unit object $I$ that are associative and unital up to coherent equivalence. A \define{braided} monoidal bicategory comes with a pseudonatural equivalence $\beta_{a,b}\maps a\ot b\to b\ot a$ and appropriate invertible modifications; it is \define{sylleptic} if there is an appropriate invertible modification $1_{a\ot b}\Rrightarrow\beta_{b,a}\circ\beta_{a,b}$ and it is \define{symmetric} if one further axiom holds. A \define{lax monoidal} pseudofunctor between monoidal bicategories $F\maps\bicat{A}\to\bicat{B}$ is a pseudofunctor equipped with pseudonatural transformations with components $\mu_{a,b}\maps Fa\otimes Fb\to F(a\otimes b)$ and $\mu_0\maps I\to FI$ along with coherent invertible modifications for associativity and unitality. This is commonly also called `weak monoidal' pseudofunctor. A \define{braided lax monoidal} pseudofunctor between braided monoidal bicategories comes with an appropriate invertible modification with components $F(\beta_{a,b})\circ \mu_{a,b}\cong\mu_{b,a}\circ \beta_{Fa,Fb}$; a \define{sylleptic lax monoidal} pseudofunctor satisfies one extra condition and a \define{symmetric lax monoidal} pseudofunctor between symmetric monoidal bicategories is just a sylleptic one. RIGHT???
